---
title: "Linear regression in R"
author: "Erin Shellman"
date: "April 13 & 20, 2015"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    theme: readable
    toc: yes
    toc_depth: 3
---

```{r setup, include = FALSE}
#knitr::opts_chunk$set(cache = TRUE)
require(dplyr)
require(ggplot2)
require(GGally)
require(scales)
require(lubridate)
require(caret)

setwd('~/projects/BI-TECH-CP303/projects/project 1')
usage = read.delim('./data/usage_2012.tsv',
                   sep = '\t',
                   header = TRUE)

stations = read.delim('./data/stations.tsv',
                   sep = '\t',
                   header = TRUE)

weather = read.delim('./data/daily_weather.tsv',
                   sep = '\t',
                   header = TRUE)
```

## Linear regression 

In this tutorial we'll learn:

* how to `merge` datasets
* how to fit linear regression models
* how to split data into test and train sets
* how to tune our models and select features

### Data preparation

We're working with the Capital Bikeshare again this week, so start by reading in
*usage*, *weather*, *stations*.
```{r, eval = FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)

usage = read.delim('usage_2012.tsv',
                   sep = '\t',
                   header = TRUE)

weather = read.delim('daily_weather.tsv',
                   sep = '\t',
                   header = TRUE)

stations = read.delim('stations.tsv',
                   sep = '\t',
                   header = TRUE)
```

### Merging data

We have three related datasets to work with, but we can't really get started 
until they're combined. Let's start with *usage* and *weather*.
The *usage* dataframe is at the resolution of the hour, while the *weather* data
are at the resolution of a day, so we know we're going to have to either 
duplicate or compress data to merge. I vote compress, let's summarize! 
```{r}
head(usage)
custs_per_day = 
  usage %>% 
    group_by(time_start = as.Date(time_start), station_start, cust_type) %>% 
    summarize(no_rentals = n(),
              duration_mins = mean(duration_mins, na.rm = TRUE))

head(custs_per_day)
```

Perfection, now we can merge!  What's the key?
```{r}
# make sure we have consistent date formats
custs_per_day$time_start = ymd(custs_per_day$time_start)
weather$date = ymd(weather$date)

# then merge. see ?merge for more details about the function
weather_rentals = merge(custs_per_day, weather, 
                        by.x = 'time_start', by.y = 'date')

# check dimensions after to make sure they are what you expect
dim(custs_per_day)
dim(weather)
dim(weather_rentals)

head(weather_rentals)
```

Great, now we want to merge on the last dataset, *stations*. What is the key to 
link *weather_rentals* with *stations*?
```{r}
final_data = merge(weather_rentals, stations, 
                   by.x = 'station_start', by.y = 'station')
dim(final_data)
dim(weather_rentals)

head(final_data[, 1:30])

# probably want to save this now!
write.table(final_data, 
            'bikeshare_modeling_data.tsv', 
            row.names = FALSE, sep = '\t')

# rename to something more convenient and remove from memory
data = final_data
rm(final_data)
```

### The `lm()` function

The function for creating a linear model in R is `lm()` and the primary 
arguments are *formula* and *data*. Formulas in R are a little funny,
instead of an = sign, they are expressed with a ~. Let's fit the model we saw in
the lecture notes: $rentals = \beta_0 + \beta_1*crossing$. There's a little snag 
we have to take care of first. Right now we've got repeated measures *i.e.* 
one measurement per day, so we need to aggregate again this time over date.
```{r}
rentals_crossing = 
  data %>% 
    group_by(station_start) %>% 
    summarize(mean_rentals = mean(no_rentals),
              crossing = mean(crossing))

head(rentals_crossing)

# plot it
ggplot(rentals_crossing, aes(x = crossing, y = mean_rentals)) +
  geom_smooth(method = 'lm', size = 2) +
  geom_point(size = 4, alpha = 0.60) +
  theme_minimal()

model = lm(mean_rentals ~ crossing, data = rentals_crossing)

# view what is returned in the lm object
attributes(model)

# get model output
summary(model)

# print model diagnostics
par(mfrow = c(2, 2))
plot(model)
```

The `attributes()` function can be called on just about any object in R and it
returns a list of all the things inside. It's a great way to explore 
objects and see what values are contained inside that could be used in other 
analysis. For example, extracting the residuals via `model$residuals` is useful
if we want to print diagnostic plots like those above.

When we run `summary()` on the `lm` object, we see the results. The *Call*
section just prints back the model specification, and the *Residuals* section
contains a summary of the distribution of the errors. The fun stuff is in the
*Coefficients* section. In the first row contains the covariate names followed 
by their estimates, standard errors, t- and p-values. Our model ends up being 
`rentals = 15 + 0.24(crosswalks)` which means that the average number of rentals
when there are no crosswalks is 15, and the average increases by 1 rental for
every four additional crosswalks.

We can fit regressions with multiple covariates the same way:
```{r}
# lets include windspeed this time
rentals_multi = 
  data %>% 
    group_by(station_start) %>% 
    summarize(mean_rentals = mean(no_rentals),
              crossing = mean(crossing),
              windspeed = mean(windspeed))

head(rentals_multi)

ggplot(rentals_multi, aes(x = windspeed, y = mean_rentals)) +
  geom_smooth(method = 'lm', size = 2) +
  geom_point(size = 4, alpha = 0.60) +
  theme_minimal()

model = lm(mean_rentals ~ crossing + windspeed, data = rentals_multi)
summary(model)
```

The model coefficients changed quite a lot when we added in wind speed. The 
intercept is now negative, and the wind speed coefficient is huge! When 
interpreting coefficients, it's important to keep the scale in mind. Wind speed 
ranges from 0.05 to 0.44 so when you multiply 1172 by 0.05 for example, you end 
up with about 60, which is within the range we'd expect.

Let's try one more, this time we'll include a factor variable.
```{r}
rentals_multi = 
  data %>% 
    group_by(station_start, is_work_day) %>% 
    summarize(mean_rentals = mean(no_rentals),
              crossing = mean(crossing),
              windspeed = mean(windspeed))

head(rentals_multi)

# plot crossings, colored by is_work_day
ggplot(rentals_multi, 
       aes(x = crossing, y = mean_rentals, color = factor(is_work_day))) +
  geom_smooth(method = 'lm', size = 2) +
  geom_point(size = 4, alpha = 0.60) +
  theme_minimal()

# plot windspeed, colored by is_work_day
ggplot(rentals_multi, 
       aes(x = windspeed, y = mean_rentals, color = factor(is_work_day))) +
  geom_smooth(method = 'lm', size = 2) +
  geom_point(size = 4, alpha = 0.60) +
  theme_minimal()

model = lm(mean_rentals ~ crossing + windspeed + factor(is_work_day), 
           data = rentals_multi)
summary(model)
```

The output looks a little funny now. There's a term called 
`factor(is_work_day)1`, what does that mean? Factors are category variables and 
their interpretation is relative to a baseline. Our factor `is_work_day` 
only has two levels, 0 and 1, and R sets 0 to the baseline by default. So the 
interpretation of that term is that we can expect about 10 additional rentals 
when it is a work day (*i.e.* `is_work_day == 0`) and the other variables are
fixed.

## The *caret* package

We'll be using the *caret* package (short for *c*lassification *a*nd 
*re*gression *t*raining) for model development because it integrates
many modeling packages in R into one unified syntax. That means more reusable
code for us! *caret* contains helper functions that provide a unified 
framework for data cleaning/splitting, model training, and comparison. I highly
recommend the 
[optional reading](https://github.com/erinshellman/BI-TECH-CP303/blob/master/reading/regression/v28i05.pdf)
this week which provides a great overview of the *caret* package.

```{r, eval = FALSE}
install.packages('caret', dependencies = TRUE)
library(caret)

set.seed(1234) # set a seed
```

Setting a seed in R insures that you get identical results each time you run
your code. Since resampling methods are inherently probabilistic, every time we 
rerun them we'll get slightly different answers. Setting the seed to the same 
number insures that we get identical randomness each time the code is run, and
that's helpful for debugging.

### Train and test data

Before any analysis in this class we'll need to divide our data into train and 
test sets. Check out 
[this](http://topepo.github.io/caret/training.html) nice overview for more 
details. The *training* set is typically about 75% of the data and is used for 
all the model development. Once we have a model we're satified with, we use our
*testing* set, the other 25% to generate model predictions. Splitting the data
into the two groups, train and test, generates two types of errors, in-sample 
and out-of-sample errors. *In-sample* errors are the errors derived from same 
data the model was built with. *Out-of-sample* errors are derived from measuring 
the error on a fresh data set. We are interested in the out-of-sample error 
because this quantity represents how'd we'd expect the model to perform in the
future on brand new data.

Here's how to split the data with *caret*:
```{r}
# select the training observations
in_train = createDataPartition(y = rentals_multi$mean_rentals,
                                   p = 0.75, # 75% in train, 25% in test
                                   list = FALSE)
head(in_train) # row indices of observations in the training set

train = rentals_multi[in_train, ]
test = rentals_multi[-in_train, ]

dim(train)
dim(test)
```

Note: I recommend doing all data processing and aggregation steps *before* 
splitting out your train/test sets.

### Training

Our workhorse function in the *caret* package in the `train` function. This
function can be used to evaluate performance parameters, choose optimal models 
based on the values of those parameters, and estimate model performance. For 
regression we can use it in place of the `lm()` function. Here's our last
regression model using the train function.

```{r}
model_fit = train(mean_rentals ~ crossing + windspeed + factor(is_work_day), 
                  data = train, 
                  method = 'lm',
                  metric = 'RMSE') 
print(model_fit)
summary(model_fit)

# get predictions
out_of_sample_predictions = predict(model_fit, newdata = test)

# compare predictions against the observed values
errors = data.frame(predicted = out_of_sample_predictions,
                    observed = test$mean_rentals,
                    error = out_of_sample_predictions - test$mean_rentals)

# eh, not so good
ggplot(data = errors, aes(x = predicted, y = observed)) + 
  geom_abline(aes(intercept = 0, slope = 1), 
              size = 3, alpha = 0.70, color = 'red') +
  geom_point(size = 3, alpha = 0.80) +
  ggtitle('out-of-sample errors') +
  theme_minimal()
```

Our prediction accuracy is not so great for this model. The RMSE is about 15
which means that on average the predictions are off by about 15 rentals.

## Parameter tuning

## Feature Selection

## Which model is the best?

Typically adding more predictors to a model will increase the $R^2$, so using
that criteria alone will cause you to favor larger models.

## Project tips

We saw how to merge the datasets together into one, but it often makes sense to 
do some aggregation before merging. For example, since we know *usage* needs to 
be aggregated and summarized to remove the date variable, it makes 
sense to merge *usage* with the weather data and summarized before merging on 
the station data.  For example:
```{r}
# we made this data frame in the merging section above
weather_rentals = merge(custs_per_day, weather, 
                        by.x = 'time_start', by.y = 'date')

# group_by all the factors and summarize the continuous variables to generate
# a final data frame that can be merged by station.
model_data = 
  weather_rentals %>% 
    group_by(
      station_start, 
      cust_type,
      weekday, 
      season_code, 
      is_holiday, 
      is_work_day, 
      weather_code) %>% 
    summarize(
      rentals = mean(no_rentals),
      duration = mean(duration_mins),
      temp = mean(temp),
      subjective_temp = mean(subjective_temp),
      humidity = mean(humidity),
      windspeed = mean(windspeed))

head(model_data)
   
# now merge on stations
final_data = merge(model_data, stations, 
  by.x = 'station_start',
  by.y = 'station')
  
data = final_data
rm(final_data)

# remove variables from the data that won't be used for modeling, e.g. lat/long
data_to_model = 
  data %>%
    select(-station_start, -id, -terminal_name, -lat, -long)

dim(data_to_model)
head(data_to_model)

model = lm(rentals ~ ., data = data_to_model)
summary(model)

# hmm, we have some weirdness in there, some stations features don't exist 
# around any of our stations, e.g. 'turning_loop'
table(data_to_model$turning_loop)

# lets remove those using the handly 'colSums' and 'which' functions
colSums(data_to_model[ , 15:143])

# we want to know 'which' columns have a sum of 0
columns_to_remove = names(which(colSums(data_to_model[ , 15:143]) == 0))

# now combine that with filter to remove those from our data
data_to_model = data_to_model[ , !(names(data_to_model) %in% columns_to_remove)]

# try the model again
model = lm(rentals ~ ., data = data_to_model)  
summary(model)

# definintely better, but we still have some weird NAs, lets troubleshoot those
table(data_to_model$vending_machine)
table(data_to_model$storage)
table(data_to_model$dojo)
table(data_to_model$tax_service)
table(data_to_model$telephone)

# all the landmarks have at most 1 in the area, so there are not enough 
# observations for least square to fit the model.
# these variables won't be helpful in prediction, so lets remove them.

data_to_model = 
  data_to_model %>% 
  select(
    -vending_machine, 
    -storage,
    -dojo,
    -tax_service,
    -telephone)
    
# try the model again
model = lm(rentals ~ ., data = data_to_model)  
summary(model)

# one last modification. our categorical variables are being treated like 
# they're continuous. lets create some factors
data_to_model$weekday = factor(data_to_model$weekday, 
                               labels = 0:6, 
                               levels = 0:6)
data_to_model$season_code = factor(data_to_model$season_code)
data_to_model$is_holiday = factor(data_to_model$is_holiday)
data_to_model$is_work_day = factor(data_to_model$is_work_day)
data_to_model$weather_code = factor(data_to_model$weather_code)

# try the model again
model = lm(rentals ~ ., data = data_to_model)  
summary(model)

# now 'is_work_day1' is NA, what gives?! remember the assumptions of linear 
# regression. our covariates must be independent - that is, not correlated. in
# this case if you know the values of weekday, you know the value of 
# is_work_day so that assumption doesn't hold. get rid of it!
data_to_model$is_work_day = NULL

# try the model again
model = lm(rentals ~ ., data = data_to_model)  
summary(model)

# ok, we've successfully hit a model but boy does it have a lot of predictors
# lets start evaluating the predictive accuracy

# select the training observations
in_train = createDataPartition(y = data_to_model$rentals,
                                   p = 0.75,
                                   list = FALSE)

train = data_to_model[in_train, ]
test = data_to_model[-in_train, ]

# when we train with the lm function, we get the same results as using lm()
model_fit = train(rentals ~ ., 
                  data = train, 
                  method = 'lm',
                  metric = 'RMSE') 

# view the relative importance of the predictors
plot(varImp(model_fit), top = 20)

rentals_predicted = predict(model_fit, newdata = test)

prediction = data.frame(rentals_predicted,
                        rentals_observed = test$rentals,
                        error = rentals_predicted - test$rentals)
summary(prediction$error)

ggplot(data = prediction, aes(x = rentals_predicted, y = rentals_observed)) + 
  geom_point(size = 3, alpha = 0.80) +
  geom_abline(aes(intercept = 0, slope = 1), 
              size = 3, alpha = 0.70, color = 'red') +
  geom_smooth() +
  theme_minimal()

ggplot(data = filter(prediction, rentals_predicted > 0), aes(x = rentals_predicted, y = rentals_observed)) + 
  geom_point(size = 3, alpha = 0.80) +
  geom_abline(aes(intercept = 0, slope = 1), 
              size = 3, alpha = 0.70, color = 'red') +
  geom_smooth() +
  theme_minimal()

ggplot(data = filter(prediction, rentals_predicted > 0), aes(x = rentals_predicted, y = error)) + 
  geom_point(size = 3, alpha = 0.80) +
  geom_smooth() +
  theme_minimal()
```

