---
title: "Association Rule Mining in R"
author: "Erin Shellman"
date: "May 16, 2015"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    theme: readable
    toc: yes
    toc_depth: 3
---

```{r setup, include = FALSE}
# my set-up
knitr::opts_chunk$set(cache = TRUE)
require(dplyr)
require(ggplot2)
require(GGally)
require(scales)
require(lubridate)
require(arules)
setwd('~/projects/BI-TECH-CP303/projects/project 3')
```


pairs = read.delim('./data/amazon0302.txt',
                   sep = '\t',
                   header = FALSE)
colnames(pairs) = c('product1', 'product2')
pairs$product1 = factor(pairs$product1)
pairs$product2 = factor(pairs$product2)

catalog = read.delim('./data/sample_catalog.tsv',
                   sep = '\t',
                   header = TRUE)
catalog = unique(catalog)                

p = merge(pairs, catalog, by.x = 'product1', by.y = 'id')
colnames(p) = c("product1", "product2", "avg_rating_1", "categories_1", 
"downloaded_1", "group_1", "reviews_count_1", "salesrank_1", "title_1")
p = merge(p, catalog, by.x = 'product2', by.y = 'id')
colnames(p) = c("product1", "product2", "avg_rating_1", "categories_1", 
"downloaded_1", "group_1", "reviews_count_1", "salesrank_1", "title_1",
"avg_rating_2", "categories_2", "downloaded_2", "group_2", "reviews_count_2", 
"salesrank_2", "title_2")

p = arrange(p, product1, product2)

```{r}
library(arules)
library(arulesViz)

trans = read.transactions("./data/people_who_bought", 
                          format = "basket", 
                          sep = ',')
itemInfo(trans) = data.frame(labels = myLabels, level1 = myLevel1)

inspect(trans[1:5])
summary(trans)

itemFrequencyPlot(trans, topN = 25)
basket_rules = apriori(trans,
                       parameter = list(sup = 0.0001, conf = 0.0001, target = 'rules'))
inspect(basket_rules)




```



txn = as(pairs, 'transactions')




