---
title: "Association Rule Mining in R"
author: "Erin Shellman"
date: "May 18, 2015"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    theme: readable
    toc: yes
    toc_depth: 3
---

```{r setup, include = FALSE}
# my set-up
knitr::opts_chunk$set(cache = TRUE)
require(dplyr)
require(ggplot2)
require(scales)
require(arules)
require(arulesViz)
setwd('~/projects/BI-TECH-CP303/projects/project 3')
bought = read.transactions('./data/people_who_bought.txt', 
                           format = 'basket', 
                           sep = ',')
catalog = read.delim('./data/product_catalog.tsv',
                     header = TRUE,
                     sep = '\t',
                     quote = '')
itemInfo(bought) = catalog
```

## Reading in the data

The *arules* package reads in data slightly differently than other packages. 
Specifically, it has its own `read.transactions` function that can be used to 
read columns stored as transactions. 

```{r, eval = FALSE}
install.packages('arules', dependencies = TRUE)
library(arules)

# read in the co-purchased items
bought = read.transactions('people_who_bought.txt', 
                           format = 'basket', 
                           sep = ',')

# read in the product catalog
catalog = read.delim('product_catalog.tsv',
                     header = TRUE,
                     sep = '\t'
                     quote = '')
```

The file *people_who_bought.txt* consists of rows of co-purchased items, in the 
form:

| 0,1,2,3,4,5
| 1,0,2,4,5,15
| 2,0,11,12,13,14
| 3,63,64,65,66,67

Where each row represents a "transaction." In reality each row corresponds to a 
product and the set of items that are frequently purchased together with the
item.

Once we've read in the itemsets and the catalog, we can append the catalog data
to the itemsets. This allows us to do convenient things like aggregate at the 
product category level.
```{r, warning = FALSE}
# attach the catalog data onto the item-sets 
itemInfo(bought) = catalog

# view the itemsets
inspect(bought[1:5])
summary(bought)

# plot the most frequent items
itemFrequencyPlot(bought, topN = 25)
```

## Exploratory data analysis

We have a new dataset again, so let's explore a bit.

```{r, warning = FALSE}
# what does the distribution of salesrank look like?
ggplot(catalog, aes(x = salesrank)) + 
  geom_histogram() 

# how about by product group?
ggplot(catalog, aes(x = salesrank)) + 
  geom_histogram() +
  facet_wrap(~ group)

# how about average rating?
ggplot(na.omit(catalog), aes(x = factor(avg_rating))) + 
  geom_bar() 

# how many items in each product group?
ggplot(na.omit(catalog), aes(x = factor(group))) + 
  geom_bar()
```

## Generating rules

In this data set we have a large catalog and relatively few transactions, which
means for any given item pairing the support will be relatively small (in this 
case support = 0.0001).

```{r}
# run the apriori algorithm
rules = apriori(bought, 
                parameter = list(sup = 0.0001, conf = 0.0001, target = 'rules'))
# view the rules
inspect(head(rules))

# sort the rules by lift
inspect(head(sort(rules, by = 'lift'), 10))

# filter by LHS to pull out specific product ids
rules_subset = subset(rules, (lhs %in% c('241')))
inspect(rules_subset)

# aggregate the rules over the product type
group_rules = aggregate(rules, itemInfo(bought)$group)
inspect(group_rules)

# quality
quality(rules) = cbind(quality(rules), coverage = coverage(rules))
```

Cool, now we have 2,601 rules to work with, but inspecting all of them manually 
sounds like a nightmare. Lucky for us there's a second package called 
*arulesViz* that gives us lots of great visualization support.

```{r, eval = FALSE}
# arules plot template
plot(x, 
     method = NULL, 
     measure = 'support', 
     shading = 'lift', 
     interactive = FALSE, 
     data,
     control = ...)
```
where:

  * *x*: is the set of rules to be visualized
  * *method*: the visualization method
  * *measure*: and shading contain the interest measures used by the plot
  * *interactive*: indicates whether you want to interactively explore 
  * *data*: can contain the transaction data set used to mine the rules 
  * *control*: list with further control arguments to customize the plot

```{r, eval = FALSE}
install.packages('arulesViz', dependencies = TRUE)

# you might need to install Rgraphviz from this repository
source('http://bioconductor.org/biocLite.R')
biocLite('Rgraphviz')

library(arulesViz)
```

We can start with a visualization of association rules as a scatter plot with 
two measures on the axes. The default `plot()` for association rules 
is a scatter plot using support and confidence on the axes. Lift is used as the 
color of the points.

```{r}
plot(rules)
head(quality(rules))

# there's also an interactive mode
# plot(rules, interactive = TRUE)
```
 
Another version of the scatter plot called two-key plot. Here support and 
confidence are used for the x and y-axes and the color of the points indicates 
the 'order,' *i.e.*, the number of items contained in the rule:
```{r}
plot(rules, shading = 'order')
```

```{r}
subrules = rules[quality(rules)$confidence > 0.5]
plot(subrules, 
     method = 'matrix', 
     measure = 'lift')

# reorder based on
plot(subrules, 
     method = 'matrix', 
     measure = 'lift', 
     control = list(reorder = TRUE))

plot(subrules, 
     method = 'matrix3D', 
     measure = 'lift', 
     control = list(reorder = TRUE))

# plot a graph
plot(subrules, method = 'graph')

# parallel coordinates plot
plot(subrules, method = 'paracoord', control = list(reorder = TRUE))
```