---
title: "Cluster Analysis in R"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    theme: readable
    toc: yes
    toc_depth: 3
date: "June 6, 2016"
---

```{r setup, include = FALSE}
# my set-up
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE,
                      fig.align = 'center')
 
options(scipen = 999)

require(dplyr)
require(ggplot2)

colleges = read.delim('./data/colleges.tsv',
                      sep = '\t',
                      header = TRUE)
```

# Read in data

We're continuing our exploration of college features, so read in the College
Scorecard data:

```{r, eval = FALSE}
require(dplyr)
require(ggplot2)

colleges = read.delim('./data/colleges.tsv',
                      sep = '\t',
                      header = TRUE)
```

We're also going to use a couple new libraries, `stringr` and `ggdendro`, so
go ahead and install those now. You can use the `packages` menu, or the following
commands:

```{r, eval = FALSE}
install.packages('stringr', dependencies = TRUE)
install.packages('ggdendro', dependencies = TRUE)
```

Let's use a realistic application that we might use clustering to solve. Say that
I work for a granting agency, like the Gates Foundation, and I want to identify
colleges that have high numbers of low-income, and first generation college 
attendees because I want to give those colleges additional funding. How should I
identify which schools to fund?

# $k$-means clustering

One approach would be to use $k$-means to identify clusters of schools meeting 
our granting requirements. 

```{r}  
college_features = 
  colleges %>%
    select(institution_name, first_gen_share, poverty_rate, 
           family_income_median, median_earnings) %>% 
    na.omit() %>%
    distinct()
  
# run k-means clustering
kmeans_cluster = kmeans(select(college_features, -institution_name), 3)

# check what attributes are in the kmeans object
attributes(kmeans_cluster)

# Find which cluster the observations belong to
head(kmeans_cluster$cluster, 10)

# centers
kmeans_cluster$centers
```

Now we can plot the clusters and how they relate to two of the variables. Note that the clusters are labeled by numbers as seen in the figure legend.
```{r}
# plot 4 clusters
ggplot(college_features, 
       aes(x = family_income_median, 
           y = median_earnings, 
           color = factor(kmeans_cluster$cluster))) +
  geom_point(alpha = 0.50) +
  theme_minimal() 
```

We can use our trusty friend `ggpairs` to visualize the resulting clusters:

```{r}
require(GGally)

college_features =
  college_features %>%
    mutate(cluster = factor(kmeans_cluster$cluster))

ggpairs(college_features, 
        lower = list(mapping = aes(color = cluster, alpha = 0.20)),
        diag = list(mapping = aes(fill = cluster, color = cluster, alpha = 0.50)),
        upper = list(mapping = aes(group = cluster)),
        columns = c('first_gen_share', 'poverty_rate', 'family_income_median', 'median_earnings'))
```

What schools should we give money to? (Use the earlier plot to figure out which cluster number to select.)

```{r}
grant_candidates = 
  college_features %>%
    filter(cluster == 1)
```

# Hierarchical Clustering

Can we find similarities in the colleges based on the composition of majors?

```{r, fig.width = 12, fig.height = 12}
require(stringr)
require(ggdendro)

# take a sample to readablility
sample = colleges[sample(nrow(colleges), 50), ]

# select all the columns that contain the string "_major_perc"
majors = sample[, str_detect(names(sample), '_major_perc')]

# put the institution name back on
majors$institution_name = sample$institution_name

# remove missing values
majors = na.omit(majors)

# compute the euclidean distance
euclidean = dist(select(majors, -institution_name), 
                 method = 'euclidean')

attributes(euclidean)

# hierarchical clustering
hier = hclust(euclidean)

attributes(hier)

# label by id
hier$labels
hier$labels = majors$institution_name

# plot dendrogram
ggdendrogram(hier, rotate = FALSE, size = 2)
ggdendrogram(hier, rotate = TRUE, size = 2)
```

Do the schools we identified as candidates for grants cluster by area of study?

```{r}
# extract the dendrogram data
dendro_data = dendro_data(hier)

attributes(dendro_data)

dendro_data$labels = unique(merge(dendro_data$labels, 
                           select(college_features, institution_name, cluster),
                           by.x = 'label',
                           by.y = 'institution_name',
                           all.x = TRUE))

ggplot(segment(dendro_data)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = label(dendro_data),
            aes(label = label, x = x, y = 0, hjust = 0, color = cluster),
            size = 2) +
  coord_flip() +
  scale_y_reverse(expand = c(0.25, 0)) +
  theme_minimal() + 
  theme(legend.position = 'bottom')
```