---
title: "Cluster Analysis in R"
author: "Erin Shellman"
date: "June 1, 2015"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    theme: readable
    toc: yes
    toc_depth: 3
---

```{r setup, include = FALSE}
# my set-up
knitr::opts_chunk$set(cache = TRUE)
require(dplyr)
require(ggplot2)
setwd('~/projects/BI-TECH-CP303/projects/project 3')
catalog = read.delim('./data/product_catalog.tsv',
                     header = TRUE,
                     sep = '\t',
                     quote = '')
```

# Introduction

This week we'll explore the product catalog a little more closely. We can 
explore product similarities using clustering and potentially use the resulting
clusters as a basis for product recommendation.

# Read in data

Just read in the product catalog this time:

```{r}
library(dplyr)
library(ggplot2)

# read in the product catalog
catalog = read.delim('product_catalog.tsv',
                     header = TRUE,
                     sep = '\t',
                     quote = '')

menace = catalog[catalog$labels == 5335, ]
data_mining = catalog[catalog$labels == 22025, ]
```

# Preprocessing

Most clustering techniques work best when the data are centered and scaled.
Recall that a variable is centered when you subtract its mean from each 
observation. A variable is scaled when you divide each observation by 
the variable standard deviation. When we center and scale a variable, the 
resulting variable follows a z-distribution with mean = 0 and sd = 1.

```{r}
# set the seed
set.seed(100)

# remove rows containing missing data
catalog = na.omit(catalog)

# most products are books, music, video or DVDs, so lets focus on those
table(catalog$group)

# subset
sub = 
  catalog %>%
    filter(group == 'Book' | group == 'DVD' | 
           group == 'Music' | group == 'Video') %>% 
    select(labels, avg_rating, downloaded, reviews_count, salesrank, group)

head(sub)
```

Before we can scale the data, we need to convert the product group from a factor
into a numeric value:

```{r}
# convert the group into dummy variables
group_dummies = model.matrix(~ factor(sub$group) - 1)
colnames(group_dummies) = c('book', 'dvd', 'music', 'video')

# append the dummy variables back on and drop the 'group' variable
sub = cbind(sub, group_dummies)
sub$group = NULL

# scale the data 
scaled = as.data.frame(scale(select(sub, -labels)))
scaled$labels = sub$labels
```

# K-means clustering

We have 4 product types, so maybe K = 4 is a good place to start?

```{r}  
# run k-means clustering
kmeans_cluster = kmeans(select(scaled, -labels), 4)

# check what attributes are in the kmeans object
attributes(kmeans_cluster)

# Find which cluster the observations belong to
head(kmeans_cluster$cluster, 10)

# centers
kmeans_cluster$centers

# plot 4 clusters
ggplot(sub, 
       aes(x = reviews_count, 
           y = salesrank, 
           color = factor(kmeans_cluster$cluster))) +
  geom_point(alpha = 0.50) +
  theme_minimal() 
```

# Hierarchical Clustering

```{r}
# Don't be a menace
menace = scaled[scaled$labels == 5335, ]

# subset to videos only
videos = 
  scaled %>%
    filter(sub$video == 1) %>%
    sample_n(30)
videos = unique(rbind(menace, videos))

# compute the euclidean distance
euclidean = dist(select(videos, -labels), method = 'euclidean')

# attributes
attributes(euclidean)

# hierarchical clustering
hier = hclust(euclidean)

# label by id
hier$labels = videos$labels

# plot dendrogram
plot(hier)

# look up nearby videos
catalog[catalog$labels == 35161, ]
catalog[catalog$labels == 58851, ]
catalog[catalog$labels == 72902, ]
```

