---
title: "Introduction to R"
author: "Erin Shellman"
date: "April 6, 2015"
output: html_document
---

## Preparation 

Exploratory data analysis is a crucial first step in any task involving data, 
and it's a perfect way to get acquainted with our tools.  This notebook is a 
guide for helping you get started using R.  It is incomplete, but I will try
to introduce new concepts and functions as they become needed.  For those 
who are new to R, new to working with data, new to programming, or all three 
please keep in mind:

> Whenever you’re learning a new tool, for a long time you’re going to suck… 
> But the good news is that is typical, that’s something that happens to everyone, 
> and it’s only temporary.

That's advice from [Hadley Wickham](http://www.r-bloggers.com/hadley-wickhams-dplyr-tutorial-at-user-2014-part-1/), 
a statistician and software developer who has revolutionized the way that people 
analyze data with R. We're going to learn to use many of his packages in this 
class, you'll be a pro at R in no time.

#### Loading libraries

One of the greatest strengths of R is the active community.  We're going to use
a handful of packages (or libraries) in this class.  Today we'll learn about
two, `dplyr` and `ggplot2`.

```{r, message = FALSE}
# Install the packages if you don't have them already.
#install.packages('dplyr')
#install.packages('ggplot2')
#install.package('GGally')

# Then load them up.
library(dplyr)
library(ggplot2)
```

#### Loading data

To start getting acquainted, we'll use one of R's many 
[preloaded datasets](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html) 
called `diamonds`. It's a dataset containing the prices and other attributes of 
about 54,000 diamonds. Loading it up for use is super simple.

```{r}
help(diamonds)
data(diamonds)
```

Most of the time we won't be using R's built-in data sets, we want to work 
with our own!  There're lots of ways to load data into R, but in this class 
you can get away with just this:

```{r, eval = FALSE}
# Set you working directory, which is normally where you data are.
setwd('path/to/your/data')
data = read.delim('data.txt',
                  header = TRUE, 
                  sep = '\t')
```

Type `?read.delim` to learn what the `header` and `sep` arguments do.

Once data are loaded, there are tons of ways to spot-check it.

```{r} 
dim(diamonds)    # print dimensions
names(diamonds)
str(diamonds)
head(diamonds, 3)
tail(diamonds, 3)
summary(diamonds)
```

## R Basics

R has many data types including lists, matrices, tables and dataframes.  We'll 
be working primarily with dataframes in this class.  Dataframes should be 
familiar and can be thought of as a spreadsheet view were columns are typically
your variable and rows are observations.

#### Variable types
* factors are categorical
* numeric 

Factors are going to be essential for plotting, and we'll talk about them in 
more detail later on.

Vectors are one-dimensional data types that contain data of the same type. 
For example `age = c(18, 30, 27, 59)` is a numeric vector and 
`relationship = c('sister', 'aunt', 'nephew')` is a vector of strings.

#### Getting help

You can access documentation for any function in R by typing `help(function_name)` or 
`?(function_name)`. The help files conform to a standard format so that they're easy 
to navigate. The most frequently used sections are *usage*, *arguments*, *value* and 
*examples*. The *usage* section describes how to call the function, *arguments*
lists all the values that can be set in the call, *value* describes what 
information the function call will return to you, and *examples* are typically
self-contained chunks of code that you can copy and paste into the console.

Sometimes it's faster and easier to get help with Google and I increasingly use
Google before using the built-in documentation.  Part of my typical workflow is 
typing questions literally into Google, with a few specifics about
the language of the solution I'm looking for.  For example, "R ggplot2 how to 
change legend font size" returns the following top 3 links:

1. [Cookbook for R » Legends (ggplot2)](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/) 
2. [r - increase legend font size ggplot2 - Stack Overflow](http://stackoverflow.com/questions/20407773/increase-legend-font-size-ggplot2)
3. [theme. ggplot2 0.9.2.1](http://docs.ggplot2.org/0.9.2.1/theme.html)

all of which answer the question.  The first link is worth bookmarking. 
It shows actual plots that make it easy to find an example that looks like what 
I want and see the code. The second is another phenomenal resource 
called Stackoverflow that is a public forum for asking programming questions and
getting answers.  I use this website probably over 50 times a day, and I bet you
will too.  The last link is the *ggplot2* documentation that lists all the 
arguments in the `theme()` element, one of which is `legend.text`.  If you feel 
like you're getting stuck on something don't spin your wheels, just start 
typing the problem into Google and you might be surprised at how easy it is to 
find solutions.

* http://www.statmethods.net/
* http://adv-r.had.co.nz/
* http://adv-r.had.co.nz/Style.html
* http://www.cookbook-r.com/
* http://stackoverflow.com/questions/tagged/r-faq%20

Finally, ask your classmates! In a analyst position in the real world you won't
be working alone. If you're stuck, chances are that your classmates are too.
Even if not, sometimes getting a fresh set of eyes on your code is all you need
to find little bugs. Help each other out!

#### Slicing and dicing

There are many ways to index into a dataframe.  Dataframes are indexed like
`df[row, col]`.  Leaving either the `row` or `col` element blank tells R to take
all rows, or all columns. Here's a handful of ways we can slice and dice a 
dataframe.

```{r}
# 1. grab a column by name and assign to edu
price = diamonds[15:20, 'price']
print(price)

# 2. or use the $ to grab the whole column
price = diamonds$price
head(price)

# 3. we can even combine the two to get rows 15 - 20 of the Education column
price = diamonds[15:20, ]$price
print(price)

# 4. what if we don't know the row numbers, but we have a condition? 
pricey = subset(diamonds, price > 18800)
print(pricey)

# 5. select just some of the columns with dplyr's 'select' function
library(dplyr)
sub = select(diamonds, carat, cut, price)
head(sub)

# 6. and then we can use dplyr's 'filter' to do the same subsetting we did in 3
filtered = filter(sub, carat > 0.40)
head(filtered)

# 7. what's cool about dplyr is that you can nest those 
head(
  select(
    filter(diamonds, carat > 0.40), 
  carat, cut, price)
)

# 8. or, even cooler, we can use the 'pipe' notation
filter_by_pipe = diamonds %>% filter(carat > 0.40) %>% select(carat, cut, price)
head(filter_by_pipe)

```

###### Self-check

1. How many diamonds cost less than $500?
2. How many diamonds cost less than $100?
3. How many diamonds cost at least $15,000?
4. What is the range of prices for diamonds of color D?

## Graphics

Visuals are the most compelling way to communicate results. At the 
exploratory stage, we generate numerous, relatively low-fidelity figures that
help familiarize the analyst with new data and guide subsequent analyses.
During the exploratory phase, the goal is to generate sensible figures
quickly without fretting over details, however **axes should always be labeled**.

Statistician and artist [Edward Tufte](http://www.edwardtufte.com/tufte/) has 
canonized some fundamental graphics tips to keep in mind as you create your
figures: 

* Highlight comparisons
* Show causality
* Show as much as possible. We'll explore faceting, color, shape, size as 
method to do this.
* Integrate evidence. Where appropriate, include text, numbers, images but only to 
the extent that they enhance the visualization's narrative.
* Figures **always** have labeled axes!! (ok, this one is mine.)

#### Visualization with *ggplot2*

R has three core plotting systems, *base*, *lattice* and *ggplot2*.
In this course we'll use the plotting library *ggplot2* because it 
combines the best parts of the other two plotting systems and uses a consistent 
syntax that makes plot code intuitive and reusable.

If you're familiar with R's `base` graphics then 
[qplot()](http://docs.ggplot2.org/0.9.3.1/qplot.html) will seem natural.  Feel 
free to read up on `qplot()` and its uses, but we're going to focus primarily on
the `ggplot` function for constructing graphics.

```{r, fig.width = 4, fig.height = 3, fig.cap = 'test caption'}
# quick scatterplot made with qplot
qplot(carat, price, data = diamonds)
```

Plot objects in *ggplot2* are made up of *geoms* and *aesthetics*. Geometric 
objects, or *geoms*, describe the type of plot, *e.g.* scatter or boxplot. 
*aesthetics* describe how to draw the plot *e.g.* color, size or location.

Every ggplot starts with a line like this:
  `ggplot(dataframe, aes(x = var1, y = var2, ...))` 
that maps data onto x and y dimensions for our plot, yet to be defined.

```{r, fig.width = 4, fig.height = 3}
# scatter plot
ggplot(diamonds, aes(x = carat, y = price)) +
  # data mapping
  geom_point(alpha = 0.10) + # draw a scatter plot
  # plot labeling
  scale_x_continuous('Carat') +
  scale_y_continuous('Price (USD)') + 
  # theme elements
  ggtitle('Are higher carat more expensive?') + 
  theme_minimal() + # simpler colors. how does it look if you remove it?
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())

# scatter plot
ggplot(diamonds, aes(x = carat, y = price)) +
  # data mapping
  geom_hex() + # draw a scatter plot
  # plot labeling
  scale_x_continuous('Carat') +
  scale_y_continuous('Price (USD)') + 
  # theme elements
  ggtitle('Are higher carat more expensive?') + 
  theme_minimal() + # simpler colors. how does it look if you remove it?
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

Think of *ggplot*s as being built up as layers. First we map the variables with 
the `ggplot` statement, then we tell ggplot what type of plot to make 
(e.g. scatter, histogram, bar plot, etc).  Finally, we can layer on nice labels, 
colors and annotation.  I highly recommend having this page up when plotting: 
<http://docs.ggplot2.org/>

```{r, fig.width = 4, fig.height = 3}
# histogram
ggplot(diamonds, aes(x = price)) +
  # data mapping
  geom_histogram(fill = 'steelblue') + # 
  # plot labeling
  scale_x_continuous('Price') +
  scale_y_continuous('Counts') + 
  # theme elements
  ggtitle("What's the distribution of prices?") + 
  theme_minimal() + 
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

See how much of that code we were able to reuse?

#### Aesthetics 

When you want to change a feature of you figure based on the value of another 
variable, you use aesthetics. Unlike the histogram above, where we could specify
the color explicitly in the *geom()*, when the color depends on the value of a
variable, it goes inside `aes()`.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  # data mapping
  geom_point(aes(color = factor(color)), alpha = 0.40) + # draw a scatter plot
  # plot labeling
  scale_x_continuous('Carat') +
  scale_y_continuous('Price (USD)') +
  scale_color_discrete('Color') + # legend title
  # theme elements
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

#### Factors

We used a new function called `factor` in the plot above to describe how to 
color the points. A `factor` is a variable type in R that represents 
categorical data, in this case the color of a diamond. Factor variables are 
very handy in *ggplot2* because they allow us to quickly change color, shape,
facet and many other graphical features by the value of a category. Let's try it
with facets.

#### Faceting 

Our colorful carat plot looks pretty good, but there are a lot of points laying 
on top of each other. When you want to compare plots across groups separately 
you can use a *facet* to split the figure up by category.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  # data mapping
  geom_point(alpha = 0.30) + # draw a scatter plot
  geom_smooth(method = 'lm') + 
  facet_wrap(~ cut) +
  # plot labeling
  scale_x_continuous('Carat') +
  scale_y_continuous('Price (USD)') +
  scale_color_discrete('Color') + # legend title
  # theme elements
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

Now we can view every grouping individually. Notice how we were able to add a 
smoother to the plot and `facet_wrap` elegantly applied it to each sub-plot.
How about with barplots?

```{r}
ggplot(diamonds, aes(clarity, fill = factor(cut))) + 
  geom_bar() +
  facet_wrap(~ cut) +     
  # plot labeling
  scale_x_discrete('Clarity') +
  scale_y_continuous('Count') +
  scale_fill_discrete('Cut') + # legend title
  # theme elements
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 6),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

#### Annotation

Annotation on a figure can add clarity and help the audience understand the 
message.  A lot of these features are not necessary for exploratory analyses,
but can be very helpful for communicating final results. For example, suppose 
we're interested in characterizing the joint distributions of carats and cuts.
These distributions are bumping and highly left-skewed, so it's difficult to
assertain the mean without further annotation.

```{r}
carat_summary = diamonds %>% group_by(cut) %>% summarize(carat_mean = round(mean(carat), 2))

ggplot(diamonds, aes(x = factor(cut), y = carat)) +
  geom_violin(fill = "grey80", colour = "#3366FF") +
  geom_text(data = carat_summary, aes(label = carat_mean, x = cut, y = carat_mean)) +
  geom_hline(yintercept = range(carat_summary$carat_mean), alpha = 0.4, linetype = 2) +
  coord_flip() +
  # plot labeling
  scale_x_discrete('Cut') +
  scale_y_continuous('Carat') +
  scale_fill_discrete('Cut') + # legend title
  # theme elements
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

Ignoring the quick aggregation we did for now, we used `geom_text` to write the
mean carat of each cut on top of the violin plot. Also we drew the range of those
means using `geom_hline` which lets us quickly assess the spread in means. We'll
talk about that cool aggregation step in detail in the next section.

###### Self-check

1. Spend some time exploring the [ggplot2 docs](http://docs.ggplot2.org/current/)
and try out some new figures. What did you learn from your figure?
    
## Exploratory data analysis

The *diamonds* dataset was a good warm-up, but let's dive into dataset we'll be 
working with for the first three weeks of class. These data are from Washington
D.C.'s bikeshare bike in 2012, and are split into two files. The first file 
called *usage_2012.tsv* is large and contains each rental event as well as
hourly weather information. The second file, called *stations.tsv* contains 
geographical data about the stations like lat/long as well as counts of nearby
amenities and road features. Let's go through a quick exploratory data analysis 
of the usage data keeping in mind that for the first project you'll need to:

1. formalize the business problem as a data mining process
2. *create* the outcome(s) variables
3. use *aggregation* and *summaries* to create model inputs

```{r}
setwd('~/projects/BI-TECH-CP303/projects/project 1')
usage = read.delim('./data/capital-bike-share/usage_2012_subset.tsv',
                   sep = '\t',
                   header = TRUE)

#usage_file = 'https://s3-us-west-2.amazonaws.com/bi-tech-cp303/project+1/capital-bike-share/usage_2012_subset.tsv'
#usage = read.delim(usage_file,
#                   sep = '\t',
#                   header = TRUE)

head(usage)
str(usage)

```

Spend some time looking over the data and getting a sense for what types of 
information are available. When I scan over a dataset I create small hypotheses 
in my mind and think about the data I'd need to disprove them. For example, 
I see a dummy variable called *is_holiday*, I wonder if the number of rentals is
higher on holiday? 

The data are basically ready to go (*you're welcome*), but lets use the 
*lubridate* package to convert our timestamps to date objects so that we can 
do lots convenient things like subtract dates.

```{r}
library(lubridate)

is(usage$time_start) # time_start is a factor

usage$time_start = ymd_hms(usage$time_start)
usage$time_end = ymd_hms(usage$time_end)

is(usage$time_start) # And now it's POSIXct (a date)
```

We can use `head(usage)` to start inspecting the data. We have a bike id, start 
and end times and locations, customer type and lots of data about the season and
the weather. Let's start exploring and see if we can make any preliminary 
observations.

```{r, fig.width = 4, fig.height = 3}
# How many unique bikes are there?
length(unique(usage$bike_id))

# How many unique stations are there?
length(unique(usage$station_start))

# What's the temperature distribution like?
summary(usage$temp)
ggplot(usage, aes(x = temp)) +
  geom_density() +
  theme_bw()

ggplot(usage, aes(x = humidity)) +
  geom_density() +
  theme_bw()
```

It's going to take forever to discover patterns like this. Fortunately we can 
use the `ggpairs` function in the *GGally* package to plot continuous variables 
all at once.

```{r}
library(GGally)
ggpairs(usage[ , c('duration_mins', 'temp', 'subjective_temp', 'humidity', 'windspeed')])
```

We can customize the upper and lower quadrants and the diagonal. Let's make 
scatter plots with a linear smoother on the lower half, correlation values on 
the upper half, and bar plots on the diagonal. Spend some time staring at the 
plot and make sure you understand how it's read.

```{r, message = FALSE, warning = FALSE}
ggpairs(usage[ , c('duration_mins', 'temp', 'subjective_temp', 'humidity', 'windspeed')],
          lower = list(continuous = 'smooth', params = c(color = 'steelblue', alpha = 0.10)),
          diag = list(continuous = 'bar', params = c(fill = '#AAB6A2', alpha = 0.80)), 
          upper = list(continuous = 'cor', params = list(corSize = 4)), 
        axisLabels = 'show') +
# theme settings
#theme_bw() +
theme(legend.position = 'none', 
        panel.grid.major = element_blank(), 
        axis.ticks = element_blank(),
        panel.border = element_blank())
```

Immediately we can see that there's a strong linear relationship between the
actual temperature and the subjective temperature. It would be a red flag if 
there wasn't! Similarly there is a negative relationship between humidity and 
wind speed (correlation = -0.33) suggesting that when it's more humid, it's less
windy. We also see that the duration of the bike ride is very heavily 
left-skewed so we can note that we might need a transformation when we get to 
the analysis.

We can do faceting in *ggpairs* plots just like we did with *ggplot2*. The 
figure below is facetted by the season:

```{r, message = FALSE, warning = FALSE}
ggpairs(usage[ , c('temp', 'humidity', 'windspeed', 'season')],
        lower = list(continuous = 'points', combo = 'facetdensity', params = list(alpha = 0.70)),
        diag = list(continuous = 'density', params = list(size = 0.75, alpha = 0.70)), 
        upper = list(continuous = 'cor', combo = 'facetdensity'),
        axisLabels = 'show', color = 'season')
```

Again we have marginal densities on the diagonal, broken up by season. Not 
surprisingly there's 4 distinct temperature distributions. We can also see that
there's a relationship between 

#### Aggregation and summary with *dplyr*

In any analysis we use aggregation techniques to summarize and compress data.
We can also use aggregation to create new variables, and that's something you'll
need to do in your first project. We got a preview of *dplyr* in the violin 
plot, but we were really just scratching the surface of it's capabilities.

List of [*dplyr* verbs](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html):

* `filter()` and `slice()`
* `arrange()`
* `select()` and `rename()`
* distinct()
* `mutate()` and `transmute()`
* `summarise()`
* `sample_n()` and `sample_frac()`

Ultimately, we want to predict the success of a station, how should we represent
'success?' Could try number of customers per day.

```{r}
# almost!
custs_per_day = usage %>% group_by(as.Date(time_start)) %>% summarize(no_rentals = n())
head(custs_per_day)

# yes!
custs_per_day = usage %>% group_by(time_start = as.Date(time_start), station_start) %>% summarize(no_rentals = n())
head(custs_per_day)

library(scales) # this package gives us pretty date formatting
ggplot(custs_per_day, aes(x = as.Date(time_start), y = no_rentals)) + 
  geom_line(aes(color = factor(station_start)))

# Maybe we want average daily rentals
mean_custs_per_day = custs_per_day %>% group_by(station_start) %>% summarize(mean_rentals = mean(no_rentals))

ggplot(mean_custs_per_day, aes(x = station_start, y = mean_rentals)) +
  geom_bar(stat = 'identity') +
  coord_flip()

ggplot(na.omit(mean_custs_per_day), aes(x = reorder(station_start, mean_rentals), y = mean_rentals)) +
  geom_bar(stat = 'identity') +
  coord_flip()

ggplot(na.omit(filter(mean_custs_per_day, mean_rentals > 1.25)), 
       aes(x = reorder(station_start, mean_rentals), y = mean_rentals)) +
  geom_bar(stat = 'identity') +
  coord_flip()

# Does the weather change usage patterns?
usage_weather = usage %>% 
  group_by(time_start = as.Date(time_start), station_start, weather_desc) %>% 
  summarize(no_rentals = n())

```

It does look like maybe we have enough variation in the station usage to be able
to make inferences about what makes a station successful.

###### Self-check

Use combinations of the *dplyr* verbs to create potential outcomes for your projects.

















