---
title: "Introduction to R"
author: "Erin Shellman"
date: "April 6, 2015"
output: html_document
---

## Preparation 

Exploratory data analysis is a crucial first step of any task involving data, 
and it's a perfect way to get acquainted with our tools.  This notebook is a 
guide for helping you get started using R.  Does this notebook give you every 
detail you need to run your analyses in R?  No, it is incomplete, but I will try
to introduce new concepts and functions only as they become needed.  For those 
who are new to R, new to working with data, new to programming, or all three 
please keep in mind:

> Whenever you’re learning a new tool, for a long time you’re going to suck… 
> But the good news is that is typical, that’s something that happens to everyone, 
> and it’s only temporary.

That's advice [from Hadley Wickham](http://www.r-bloggers.com/hadley-wickhams-dplyr-tutorial-at-user-2014-part-1/), 
a statistician and software developer who has revolutionized the way that people 
use R.  We're going to use a lot of his software in this class.

#### Loading libraries

One of the greatest strengths of R is the active community.  We're going to use
a handful of packages (or libraries) in this class.  Today we'll learn about
two, `dplyr` and `ggplot2`.

```{r, message = FALSE}
# Install the packages if you don't have them already.
# install.packages('dplyr')
# install.packages('ggplot2')

# Then load them up.
library(dplyr)
library(ggplot2)
```

#### Loading data

To start getting acquainted, we'll use on of R's many 
[preloaded datasets](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html) 
called `swiss`.  Loading it up for use is super simple.

```{r}
data(swiss)
```

Most of the time you won't be using R's built-in data sets, you want to work 
with your own!  There are lots of ways to load data into R, but in this class 
you can get away with just this:

```{r, eval = FALSE}
# Set you working directory, which is normally where you data are.
setwd('path/to/your/data')
data = read.delim('data.txt',
                  header = TRUE, 
                  sep = '\t')
```

Type `?read.delim` to learn what the `header` and `sep` arguments do.

Once data are loaded, there are tons of ways to spot-check it.

```{r} 
dim(swiss) # print dimensions
names(swiss)
str(swiss)
head(swiss, 3)
tail(swiss, 3)
summary(swiss)
plot(swiss)
```

## R Basics

R has many data types including lists, matrices, tables and dataframes.  We'll 
be working primarily with dataframes in this class.  Dataframes should be 
familiar and can be thought of as a spreadsheet view were columns are typically
your variable and rows are observations.

#### Variable types
* factors - categories, emphasize importance for plotting.
* numeric

Vectors are one-dimensional data types that contain data of the same type.  For example `age = c(18, 30, 27, 59)` is a numeric vector and 
`relationship = c('sister', 'aunt', 'nephew')` is a vector of strings.

#### Getting help

You can access the documentation for any function in R by typing `help(lm)` or 
`?(lm)`.  The help files conform to a standard format so that they're easy 
to navigate.  Most likely you'll make use of *usage*, *arguments*, *value* and 
*examples* sections when exploring how to use a function.

Sometimes it's faster and easier to get help with Google and I increasingly use
Google before using the built-in documentation.  Part of my typical workflow is 
typing questions literally into Google, with a few specifics about
the language of the solution I'm looking for.  For example, "R ggplot2 how to 
change legend font size" returns the following top 3 links:

1. [Cookbook for R » Legends (ggplot2)](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/) 
2. [r - increase legend font size ggplot2 - Stack Overflow](http://stackoverflow.com/questions/20407773/increase-legend-font-size-ggplot2)
3. [theme. ggplot2 0.9.2.1](http://docs.ggplot2.org/0.9.2.1/theme.html)

all of which answer the question quickly.  The first link is worth bookmarking. 
It shows actual plots that make it easy to find an example that looks like what 
I want and see the code. The second is another phenonemal resource 
called Stackoverflow that is a public forum for asking programming questions and
getting answers.  I use this website probably over 50 times a day, and I bet you
will too.  The last link is the ggplot2 documentation that lists all the 
arguments in the `theme()` element, one of which is `legend.text`.  If you feel 
like you're getting stuck on something don't spin your wheels, just start 
typing the problem into Google and you might be surprised at how easy it is to 
find solutions.

* http://www.statmethods.net/
* http://adv-r.had.co.nz/
* http://adv-r.had.co.nz/Style.html
* http://www.cookbook-r.com/
* http://stackoverflow.com/questions/tagged/r-faq%20

#### Slicing and dicing

There are many ways to index into a dataframe.  Dataframes are indexed like
`df[row, col]`.  Leaving either the `row` or `col` element blank tells R to take
all rows, or all columns.

```{r}
# grab a column by name and assign to edu
edu = swiss[15:20 , 'Education']
print(edu)
# or use the $ to grab the whole column
edu = swiss$Education
head(edu)

#subset() demo
```

## Graphics

In virtually any setting graphics are going to be the best, most compelling way
of communicating results to your customer.  At the exploratory stage of our 
analysis we're going to generate numerous, relatively low-fidelity figures to 
help us understand the domain as analysts and to guide our subsequent analyses.
During the exploratory phase, we want to be able to generate sensible figures
quickly without fretting over details, however **axes should always be labeled**.

Some graphics tips: 
* Highlight comparisons
* Show causality
* Show as much data on a single plot as possible - multivariate data 
  - can do that with faceting, color, shape, size
* Integrate evidence - text, numbers, images

Plots to choose from
* Box plots (1D)
* Histograms / Density plots (1D)
* Bar plots (1D)
* Scatter plots (2D)
* Facets (2D)

Histogram with a rug! http://docs.ggplot2.org/0.9.2.1/geom_rug.html

Things to show
* Adding lines and annotation for clarity
* Facet_wrap - discuss factors
* Fills and color
* show geom_density
* smoother
* an example of some automation of figure generation?

#### Visualization with `ggplot2`

R has three core plotting systems, `base`, `lattice` and `ggplot2`.
In this course we're going to use a plotting library called `ggplot2` because I 
think it produces the best plots, and uses a syntax that is more consistent and 
reusable than base R graphics.

If you're familiar with R's `base` graphics then 
[qplot()](http://docs.ggplot2.org/0.9.3.1/qplot.html) will seem natural.  Feel 
free to read up on `qplot()` and its uses, but we're going to focus primarily on
the `ggplot` function for constructing graphics.

```{r, fig.width = 4, fig.height = 3, fig.cap = 'test caption'}
qplot(Fertility, Education, data = swiss)
```

Plots in `ggplot2` are made up of *aesthetics* and *geoms*.

The components of a `ggplot` are
* input data
* aesthetic mappings
* geoms
* facets - optional
* stats
* scales
* coordinate system

Every ggplot starts with a line like this:
  `ggplot(dataframe, aes(x = var1, y = var2, ...))` 
that maps data onto x and y dimensions for our plot, yet to be defined.

```{r, fig.width = 4, fig.height = 3}
# scatter plot
ggplot(swiss, aes(x = Catholic, y = Fertility)) +
  # data mapping
  geom_point() + # draw a scatter plot
  # plot labeling
  scale_x_continuous('Catholic (%)') +
  scale_y_continuous('Fertility') + 
  # theme elements
  ggtitle('Are Catholics more Fertile?') + 
  theme_minimal() + # simpler colors. how does it look if you remove it?
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

Think of ggplots as being built up as layers. First we map the variables with 
the ggplot statement, then we tell ggplot what type of plot to make 
(e.g. scatter, histogram, bar plot, etc).  Finally, we can layer on nice labels, 
colors and annotation.  I highly recommend having this page up when plotting: 
<http://docs.ggplot2.org/>

```{r, fig.width = 4, fig.height = 3}
# histogram
ggplot(swiss, aes(x = Catholic)) +
  # data mapping
  geom_histogram(fill = 'steelblue') + # 
  # plot labeling
  scale_x_continuous('Catholic') +
  scale_y_continuous('Counts') + 
  # theme elements
  ggtitle("What's the distribution of Catholics?") + 
  theme_minimal() + 
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

See how much of that code we were able to reuse?

#### Aesthetics 

When you want to change a feature of you figure based on the value of another 
variable, you use aesthetics. Unlike the histogram above, where we could specify
the color explicitly in the *geom()*, when the color depends on the value of a
variable, it goes inside `aes()`.

```{r}
# bubble plot
ggplot(swiss, aes(x = Catholic, y = Fertility, size = Infant.Mortality)) +
  # data mapping
  geom_point() + # draw a scatter plot
  # plot labeling
  scale_x_continuous('Catholic (%)') +
  scale_y_continuous('Fertility') + 
  # theme elements
  ggtitle('Are Catholics more Fertile?') + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 12),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
```

#### Annotation

Annotation on a figure can add clarity and help the audience understand the 
message.  A lot of these features are not necessary for exploratory analyses,
but can be very helpful for communicating final results.

* xlab, ylab, labs, ggtitle, annotate, themes

swiss$is_catholic = ifelse(swiss$Catholic > 80, 'Catholic', 'Protestant')
table(swiss$is_catholic)

# boxplot
ggplot(swiss, aes(x = rownames(swiss), y = Fertility)) +
  # data mapping
  geom_box plot() + 
  # plot labeling
  scale_x_continuous('Catholic') +
  scale_y_continuous('Counts') + 
  # theme elements
  ggtitle("What's the distribution of Catholics?") + 
  theme_minimal() + 
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    title = element_text(size = 17),
    axis.title = element_text(size = 15),
    axis.ticks = element_blank())
    
##