---
output:
  pdf_document: default
  html_document: default
---

```{r, include = F}
library(dplyr)
library(ggplot2)

colleges = read.delim("../colleges.tsv", sep = '\t', header = TRUE)
```

# Answer Key: Unsupervised Learning II: Clustering

**K-means clustering randomizes the initial values for optimization, so results may vary everytime it is run. This answer key uses a `set.seed(10)` to get consistent results.**

**Problem statement: A granting agency wants to identify colleges that have high numbers of low-income, and first generation college attendees to give those colleges additional funding.**

1. In the clustering tutorial, we used k-means clustering to identify 3 clusters of colleges using these criteria. 
    
    **A.** Replicate this analysis using the code in the tutorial to generate those 3 clusters and append the cluster levels to the `college_features` dataset.

    ```{r, eval = F, messsage = F, warning = F, comment = F}
library(dplyr)
library(ggplot2)
    ```

    ```{r, eval = T, messsage = F, warning = F, comment = F}

### For reproducible results
set.seed(10)

### Load the colleges datasets on your machine
### colleges = read.delim("colleges.tsv", sep = '\t', header = TRUE)

college_features = colleges %>% 
  select(institution_name, first_gen_share, poverty_rate, family_income_median,
         median_earnings, top_ten) %>% 
  na.omit() %>%
  distinct()

kmeans_cluster = kmeans(select(college_features, -institution_name, -top_ten), 3)

college_features$cluster = kmeans_cluster$cluster
    ```
    
    **B.** What is the median family income for each cluster (**hint:** see `kmeans_cluster$centers` from the tutorial)?
    ```{r}
    kmeans_cluster$centers
    ```

 
    **C.** Subset the colleges_features dataset on the cluster with the lowest `family_income_median`, call this new data grant_candidates. **Note:** in the tutorial, grant_candidates were from Cluster 1, you could find that a different cluster from your analysis has the lowest `family_income_median` when you look at `kmeans_cluster$centers`.
    ```{r}
    grant_candidates = college_features %>% filter(cluster == 3)
    ```
    
    **D.**  How many universities are in the cluster of grant receivers? 
    
    ```{r}
    dim(grant_candidates)
    ```

    *2,501 are in the lowest family income cluster and would receive grants using this method.*

\clearpage

2. Upon review you’re informed that there are too many universities receiving grants. The granting agency really likes the cluster approach but suggests you make 5 clusters instead of 3.

    **A.** Redo the k-means analysis above but create 5 clusters instead of 3. **Note:** If you appended cluster onto your `college_features` dataset, make sure to remove it before redoing the k-means analysis.

    ```{r, eval = T, messsage = F, warning = F, comment = F}


### re-running this code so the cluster variable isn't included
    
college_features = colleges %>% 
  select(institution_name, first_gen_share, poverty_rate, family_income_median,
         median_earnings, top_ten) %>% 
  na.omit() %>%
  distinct()

kmeans_cluster = kmeans(select(college_features, -institution_name, -top_ten), 5)

college_features$cluster = kmeans_cluster$cluster

    ```
    
    **B.** Again subset the data on the cluster with the lowest `family_income_median`. How many universities will receive a grant now? What is the median and range of `family_income_median` of these universities and how does it compare to your answers in Question 1?

    ```{r}
### instead of printing the cluster results, you can find the minimum family income 
### using programmatic methods
grant_candidates = college_features %>% 
  filter(cluster == which.min(kmeans_cluster$centers[,"family_income_median"]))

dim(grant_candidates)
    ```

    *Compared to the 2,501 universities identified in 1D, there are only 31 that are in the cluster now.*

    **C.** You will likely find that there were two clusters out of the five with low but similar `family_income_median`. Among these two clusters, what else determined which cluster these universities were assigned to (**hint:** look at the centers again)? Based on those other variables, do you think we made the correct decision to distribute grants considering only `family_income_median`?

    ```{r}
    kmeans_cluster$centers
    ``` 

    *Despite having the lowest median family income at $14,966, our grant cluster has the highest median earnings after graduation among all of the clusters. The cluster with the second lowest median family income ($17,116) has the highest poverty rate and lowest post-graduation earnings, so would seemingly be a better target for a grant program.*


\clearpage

3. Hierarchical clustering: Part of the grant is to reformulate curriculums to better match top ten universities.    


    **A.** Subset your colleges dataset using the following code. The !is.na(sat_verbal_quartile_1) removes universities that do not have SAT admission criteria, so we are looking at similar degree-granting universities. What other criteria are we using to subset? 

    ```{r, eval = T}
    grant_colleges = 
      colleges %>% 
      filter(
        !is.na(sat_verbal_quartile_1) & family_income_median < 40000 & median_earnings < 30000
      )
    
    top_ten_schools = colleges %>% filter(top_ten == TRUE)
  
    heir_analysis_data = rbind(grant_colleges, top_ten_schools)
    ```

    *We are also selecting colleges with median family income less than \$40,000 AND median earnings after graduation less than \$30,000.*
    
    **B.** Replicate the heirarchical clustering from the tutorial comparing major percentages using `heir_analysis_data` dataset. Which universities are the most different from the top ten schools in terms of majors?
    
    ```{r, eval = T}
    
major_perc = heir_analysis_data %>% 
  select(institution_name, top_ten, contains("_major_perc")) %>%
  na.omit()

### Calculate distances
euclidean = dist(select(major_perc, -institution_name, -top_ten), method = "euclidean")

### hierarchical clustering
hier = hclust(euclidean)

### Relabel the nodes to be institution names
hier$labels = major_perc$institution_name
  
    ```

    ```{r, eval = T}
### plot using ggdendro
library(ggdendro)
ggdendrogram(hier, rotate = TRUE, size = 2)

    ```

    ```{r, eval = T, warning=F}
### plot using ggplot2

# necessary step for plotting with ggplot2
dendro_data = dendro_data(hier) 

# merges on the top_ten variable for plotting
dendro_data$labels = dendro_data$labels %>%
  left_join(select(college_features, institution_name, top_ten),
            by = c("label" = "institution_name"))

ggplot(segment(dendro_data)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_text(data = label(dendro_data), 
            aes(label = label, x = x, y = 0, hjust = 0, color = top_ten), size = 2) + 
  coord_flip() +
  scale_y_reverse(expand = c(0.25, 0)) + 
  theme_minimal() +
  theme(legend.position = "bottom")
    
    ```
    
    **C.** How else can we compare the grantee schools to the top ten schools? Explore using any of the methods we learned in this class.
    
    *Open question for exploration.*
