---
output:
  pdf_document: default
  html_document: default
---

```{r, include = F}
library(dplyr)
library(ggplot2)
library(GGally)
library(caret)

default = read.delim("../credit_card_default.tsv", sep = "\t", header = TRUE)
```

# Predicting credit card defaults with logistic regression


## Instructions

Follow the steps below to familiarize yourself with the credit card default data set and prepare it for a machine learning analysis. You’ll start by building a classifier using logistic regression and continue the model building in the next lesson.

1. Read in the data using the `read.delim` function. Then use `ggplot2`, `ggpairs`, and `dplyr` to identify interesting relationships in the data. Write a short description of one interesting pattern you identified.

*There is lots of exploratory data analysis to do. Below we show an example of exploring the numerical variables and then the pay_month variables.*

```{r, eval = F, include = F}
library(dplyr)
library(ggplot2)
library(GGally)
library(caret)

### Load the forest fires datasets on your machine by setting the working directory 
### or specify the directory of your data 

### This code is commented out and will NOT run
### default = read.delim("credit_card_default.tsv", sep = "\t", header = TRUE)
```

```{r, warning = F,message=F}
default$default_next_month = factor(default$default_next_month)

default$sex = factor(default$sex)
default$education = factor(default$education)
default$marriage = factor(default$marriage)
default$pay_sept = factor(default$pay_sept)
default$pay_aug = factor(default$pay_aug)
default$pay_july = factor(default$pay_july)
default$pay_june = factor(default$pay_june)
default$pay_may = factor(default$pay_may)
default$pay_april = factor(default$pay_april)

# example, look at numerical variables 
ggpairs(select(default, limit_bal, age, default_next_month))
```

```{r, fig.height=3, fig.width=3}
# example, look at a few pay variables 
sept_pay_default = default %>%
  group_by(pay_sept) %>%
  summarize( default_rate = sum(default_next_month == 1)/ n())

ggplot(data = sept_pay_default, aes(x = pay_sept, y =default_rate )) +
  geom_point()

aug_pay_default = default %>%
  group_by(pay_aug) %>%
  summarize( default_rate = sum(default_next_month == 1)/ n())

ggplot(data = aug_pay_default, aes(x = pay_aug, y =default_rate )) +
  geom_point()


```

\clearpage

2. Construct at least one new feature to include in model development. You might choose to create a new feature based on your findings from the exploratory data analysis. Plot or summarize the new variable and interpret the result. If your new variable is numerial, use color or a facet to show the relationship between your new feature and the outcome variable `default_next_month`.

```{r}

default$sept_debt = ifelse(default$pay_sept %in% 1:8, 1, 0)

sept_debt_rate = default %>%
  group_by(sept_debt) %>%
  summarize( default_rate = sum(default_next_month == 1)/ n())

ggplot(sept_debt_rate, aes(x = sept_debt, y = default_rate)) +
  geom_point()

```

\clearpage

3. Use the `createDataPartition` function from the `caret` package to split the data into a training and testing set. Pre-process the data with `preProcess` as needed.

```{r}

#use my sept_debt variable instead of pay_sept
default = select(default, -pay_sept)

in_train = createDataPartition(y = default$default_next_month, p = 0.8, list = FALSE)
default_train = default[in_train, ]
default_test = default[-in_train, ]

preprocess_steps = preProcess(default_train,
                              method = c("center", "scale", "nzv"))

default_train_proc = predict(preprocess_steps, default_train)

```

4. Fit at least 3 logistic regression models.

```{r,warning=F,eval=T,message=F}
full_model = train(y = default_train$default_next_month,
                   x = select(default_train, -default_next_month),
                   method = "glm", family = "binomial")

demo_model = train(y = default_train$default_next_month,
                   x = select(default_train, age, education, marriage, sex),
                   method = "glm", family = "binomial")

balance_model = train(y = default_train$default_next_month,
                   x = select(default_train, starts_with("pay"), starts_with("bill"),
                              limit_bal, sept_debt),
                   method = "glm", family = "binomial")

sept_info_model = train(y = default_train$default_next_month,
                   x = select(default_train, contains("sept")),
                   method = "glm", family = "binomial")

```

5. Use the `dotplot` function to compare the accuracy of the models you constructed in 4. Which model performed the best in terms of predictive accuracy?

*In terms of accuracy the full model performs the best. Interestingly, the kappa for the September balance information model is comparable or better than the other models. There were a lot of warnings running these models, it would be a good idea to further process the categorical variables. We could use `model.matrix` to find near zero variance in the categories or do something similar to `sept_debt` for each month.*

```{r,eval=T}
results = resamples(list(full_model = full_model,
                         demo_model = demo_model,
                         balance_model = balance_model,
                         sept_info_model = sept_info_model))
                    
dotplot(results)

```
