---
output:
  pdf_document: default
  html_document: default
---

```{r, include = F}
library(dplyr)
library(ggplot2)

ff = read.delim("../forestfires.tsv", sep = '\t', header = TRUE)

library(caret)

```


#  Creating a machine learning analysis plan

## Instructions

If you haven't already, download [`forestfires.tsv data`](https://canvas.uw.edu/courses/1105384/files/39712993) and [documentation](https://canvas.uw.edu/courses/1105384/files/39712997/download?wrap=1). Use the `caret` package to prepare the forest fire data for a machine learning analysis. When you finish preparing the data, save your script and upload it to Canvas.

```{r, eval = F}
library(caret)
library(dplyr)
library(ggplot2)

### Load the forest fires datasets on your machine by setting the working directory 
### or specify the directory of your data 

### This code is commented out and will NOT run
### ff = read.delim("forestfires.tsv", sep = '\t', header = TRUE)

```

\clearpage

1. Make at least one new feature and plot it against the burn area. From visual inspection does there appear to be a relationship between the new feature and the burn area?

```{r, fig.height=3, fig.width=3}

ff$is_weekend = ifelse(ff$day %in% c("sat", "sun"), 1, 0)
ff$is_weekend = factor(ff$is_weekend)

ff$log_area = log(ff$area + 1)

ggplot(ff, aes(x = is_weekend, y = area)) +
  geom_boxplot()

ggplot(ff, aes(x = is_weekend, y = log_area)) +
  geom_boxplot()

```

*Since the area variable is so skewed, it is easier to see relationships on the log scale (second plot). The relationship between `is_weekend` and burn area is not overwhelming but the median is lower on weekend. There is also a ggpairs plot of the `is_weekend` variable in the tutorial.*
  
  
\clearpage

2. Use `createDataPartition` to split 80% of the forest fire data into a training set.

```{r}
in_train = createDataPartition(y = ff$log_area, p = 0.8, list = FALSE)
ff_train = ff[in_train, ]
ff_test = ff[-in_train, ]
```


3. Use `preProcess` to prepare your data for analysis. What, if any, variables were removed for near zero variance? 

*One variable (`rain`) is removed. Following the tutorial, using the `model.matrix` method helps find other near zero variance categories in the factor variables.*

```{r}
preprocess_steps = preProcess(select(ff_train, FFMC, DMC, DC, ISI, temp, RH, wind, rain),
                              method = c("center", "scale", "nzv"))

preprocess_steps

nearZeroVar(ff_train, saveMetrics = TRUE)

```